Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-25 index.qmd
section:issues,CaseV; process:write

<!-- 2. Thurstone's Theory of CJ: discriminal difference -->
However, since the discriminal process of a single stimulus is not directly observable, the *law of comparative judgment* becomes essential. This law states that in pairwise comparisons, the stimulus positioned further along the trait continuum is perceived as having a higher level of that trait. Thus, the theory assumes the observed dichotomous outcome is determined by the distribution of the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*. This indicates that the outcome depends on the relative distance between stimuli, rather than their absolute positions on the trait continuum.

<!-- 3. Example-->
These concepts are more easily understood through an example. For instance, in the context of evaluating text quality, @fig-discriminal_process could depict the underlying discriminal process distributions for two written texts, highlighting differences in their discriminal dispersions and modal discriminal processes along the quality trait continuum. Furthermore, @fig-discriminal_difference could display the discriminal difference distribution for these texts, showing that text A is perceived to exhibit significantly higher quality than text B, as indicated by the shaded gray area. Consequently, the dichotomous outcome of this comparison would probably favor text A.

@@

<!-- 4. Thurstone's Cases and their simplifying assumptions  -->
Importantly, the general form of Thurstone's theory primarily addressed pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp. 267]. Thus, for practical application, Thurstone introduced five distinct cases derived from this general form, each defined by progressively simplifying assumptions. @tbl-thurstone_cases summarizes these cases, highlighting key assumptions such as the distribution of discriminal processes, the similarity of discriminal dispersions across stimuli, the correlation between stimuli, and the number of judges performing the comparisons. For a comprehensive discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248-253].

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
Despite its reliance on the largest number of simplifying assumptions [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677], Case V remains the most widely used case in the CJ literature. This popularity is largely due to its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors the assumptions of Case V, with one key difference: while Case V assumes a Normal distribution for the discriminal processes of stimuli, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions share similar statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16] (see @fig-logistic_vs_normal).

@@

<!-- 3. What does the zero correlation between stimuli imply? and does it hold? -->
The assumption of *zero correlation between stimuli* can be better understood through an example. For instance, when using pairwise comparisons to evaluate text quality, the assumption implies that a judge's perception of a trait in one text does not influence his perception of the same trait in another text. Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp. 268]. This cancellation was mathematically demonstrated by @Andrich_1978, using the BTL model under the assumption of additive biases in the discriminal processes. However, it is easy to imagine at least two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli, and when an additional hierarchical structure is relevant to the stimuli.

@@

<!-- 4. The first scenario where the assumption does not hold -->
In the first scenario, the intricate aspects of multidimensional, complex traits and heterogeneous stimuli may introduce dependencies between stimuli. Research on text quality suggests that when judges evaluate these traits, they often rely on various intricate aspects of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. In this context, it is not inconceivable that these aspects, being neither equally weighted nor opposing, may unevenly influence judges' perceptions, resulting in biases that resist cancellation. For example, this might occur when a judge assessing the argumentative quality of a text places disproportionate emphasis on grammatical accuracy, ultimately favoring texts with fewer errors but weaker arguments. While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of judges' biases, supporting the idea that the different factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 5. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections created by the additional hierarchical structure may introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010]. Nevertheless, despite recognizing such hierarchical structures in CJ data, the statistical handling of this extra source of dependency in the CJ literature has been inadequate. For instance, in cases where the CJ data included multiple samples of stimuli from the same individuals, researchers have often relied on (averaged) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. This approach, however, has the significant limitation of ignoring the uncertainty associated with the scores (refer to section @sec-theory-issue2 for a detailed discussion of this issue).

@@

<!-- 6. What does the equal dispersion between stimuli imply? and does it hold? -->
In contrast, the assumption of *equal dispersion between stimuli* suggests that the variability in the perceived trait level of the stimuli is the same across stimuli. While Thurstone acknowledged that this assumption may be violated when "dealing with less conspicuous attributes or with less homogeneous stimuli" [@Thurstone_1927a, pp. 374], no study explicitly proposes that this assumption could also be violated due to the presence of an additional hierarchical (grouping) structure relevant to the texts. One such scenario might arise, for example, when comparing texts produced by university and secondary school students. In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students, who exhibit a broader range of writing abilities, would show greater variability in the quality of their texts. Although this example is somewhat contrived, it effectively illustrates how assuming equal dispersions across texts can overlook meaningful differences in the reliability of text quality across groups or individuals.

@@

<!-- 7. But what are the challenges when these assumptions are violated? -->
["however" paragraph]{style="color:red;"}

>>>>

<!-- 2. Thurstone's Theory of CJ: discriminal difference -->
Nevertheless, the theory posits that because the discriminal process of a single stimulus is not directly observable, the *law of comparative judgment* becomes essential. This law states that in pairwise comparisons, the stimulus positioned further along the trait continuum is perceived as having a higher level of that trait. This emphasize that the outcome depends on the relative distance between stimuli, rather than their absolute positions on the trait continuum. Thus, the theory assumes the observed dichotomous outcome is determined by the distribution of the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*.

<!-- 3. Example-->
These concepts are more easily understood through an example. For instance, in the context of evaluating text quality, @fig-discriminal_process depicts the underlying discriminal process distributions for two written texts. The figure highlights differences in the texts' positions along the quality trait continuum, represented by their modal discriminal processes, as well as differences in the variability of their perceived trait levels, represented by their discriminal dispersions. Furthermore, @fig-discriminal_difference displays the discriminal difference distribution for these texts. This figure shows that text A is perceived to exhibit significantly higher quality than text B, as indicated by the shaded gray area. Consequently, the dichotomous outcome of this comparison would likely favor text A.

@@

<!-- 4. Thurstone's Cases and their simplifying assumptions  -->
Importantly, the general form of Thurstone's theory primarily addressed pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp. 267]. Thus, to facilitate its practical application, Thurstone developed five distinct cases derived from this general form. Each case progressively introduces additional simplifying assumptions related to different features of the theory, such as the distribution of discriminal processes, the uniformity of discriminal dispersions across stimuli, the correlation between stimuli, and the number of judges performing the comparisons. A summary of these cases is provided in @tbl-thurstone_cases. For a detailed examination of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248-253].

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
Despite its reliance on the largest number of simplifying assumptions [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677], Case V remains the most widely used case in the CJ literature. This popularity is largely due to its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors the assumptions of Case V, with one key difference: while Case V assumes a Normal distribution for the discriminal processes of the stimuli, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions share similar statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16] (see @fig-logistic_vs_normal).

@@

<!-- 3. What does the zero correlation between stimuli imply? and does it hold? -->
The assumption of *zero correlation between stimuli* can be better understood through an example. For instance, when using pairwise comparisons to evaluate text quality, the assumption implies that a judge's perception of a trait in one text does not influence his perception of the same trait in another text. Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp. 268]. This cancellation was mathematically demonstrated by @Andrich_1978, using the BTL model under the assumption of discriminal processes with additive biases. However, it is easy to imagine at least two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli, and when an additional hierarchical structure is relevant to the stimuli.

@@

<!-- 4. The first scenario where the assumption does not hold -->
In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between heterogeneous stimuli due to certain judges' biases that resist cancellation. Research on text quality indicates that when judges evaluate such traits, they often rely on various intricate aspects of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. In this context, it is not inconceivable that these aspects, being neither equally weighted nor opposing, may unevenly influence judges' perceptions, resulting in biases that resist cancellation. For example, this could occur when a judge assessing the argumentative quality of a text places disproportionate emphasis on grammatical accuracy, ultimately favoring texts with fewer errors but weaker arguments. While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of judges' biases, supporting the idea that the factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 5. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections created by the additional hierarchical structure may introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010]. Nevertheless, despite recognizing such hierarchical structures in CJ data, the statistical handling of this extra source of dependency in the CJ literature has been inadequate. For instance, in cases where the CJ data included multiple samples of stimuli from the same individuals, researchers have often relied on (averaged) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. This approach, however, has the significant limitation of ignoring the uncertainty associated with the BTL scores (refer to section @sec-theory-issue2 for a detailed discussion of this issue).

@@

<!-- 6. What does the equal dispersion between stimuli imply? and does it hold? -->
In contrast, the assumption of *equal dispersion between stimuli* suggests that the variability in the perceived trait level of the stimulus is the same across all stimuli. While Thurstone acknowledged that this assumption may be violated when "dealing with less conspicuous attributes or with less homogeneous stimuli" [@Thurstone_1927a, pp. 374], no study explicitly proposes that this assumption could also be violated due to the presence of an additional hierarchical (grouping) structure relevant to the texts. One such scenario might arise, for example, when comparing texts produced by university and secondary school students. In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students, who exhibit a broader range of writing abilities, would show greater variability in the quality of their texts. Although this example is somewhat contrived, it effectively illustrates how assuming equal dispersions across texts can overlook meaningful differences in the reliability of text quality across groups or individuals.

@@

<!-- 7. But what are the challenges when these assumptions are violated? -->
["However" paragraph]{style="color:red;"}
<!-- But the psychometric literature warns against ignoring additional traits that might confound the measurement of the trait of interest. Particularly, @vanderLinden_et_al_2017_I, [pp. 346] suggests that when the assumption of unidimensionality, i.e, when the model expects to measure only one trait, is violated and more than one dimension is necessary to describe the position of the stimuli in the trait space, The association between the responses to the items given one proficiency parameter does then not vanish, i.e. violating the usual conditional independence assumption -->

<!-- @Hoyle_et_al_2023 [pp. 340-341] indicates that the implication of forcing multidimensional data into a unidimensional model is that the standard measurement error for the trait become too small, due to the multidimensionality. Moreover,  -->

<!-- Suppose the assumption of unidimensionality is violated and more than one dimension is necessary to describe the test takers’ position in the latent space. The association between the responses to the items given one proficiency parameter does then not vanish, i.e. violating the usual conditional independence assumption [@vanderLinden_et_al_2017_I, pp. 346] -->

<!-- Furthermore, statistical literature also caution about ignoring clustering (grouping) structures present in the data, as these might  -->
